{\steve}: A hierarchical Bayesian model for Supernova Cosmology

### 2018-12-20 papers/methods/paper.tex

One large difference between traditional methods and BHM methods is that we treat selection effects by incorporating selection efficiency into our model, rather than relying on simulation-driven data corrections. We describe the probability that the events we observe are drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are observed and pass cuts.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As our model so far describes components of a basic likelihood $P({\rm data}|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
where the denominator represents an integral over all potential data $D$. This is derived in Appendix \ref{app:selection1} As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high-dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by Hamiltonian Monte-Carlo, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects from their full expression in all of $\theta$-space, to apparent magnitude and redshift space independently (not dependent on $x_1$ or $c$), such that the denominator of equation \eqref{eq:main}, denoted now $d$ for simplicity, is given as
\begin{equation}
d = \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz, \label{eq:w1}
\end{equation}
where $P(m_B | z, \theta)$ can be expressed by translating the underlying $M_B$, $x_1$, and $c$ population to $m_B$ given cosmological parameters. A full derivation of this can be found in Appendix \ref{app:selection2}.

@@

\subsection{General Selection Effects}
\label{app:selection1}
When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{equation}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{equation}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments that have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{equation}
\mathcal{L}(\theta; {\rm data}) = P({\rm data} | \theta, S). \label{eq:app_like}
\end{equation}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{eqnarray}
P({\rm data} | S , \theta) P(S, \theta) &=& P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &=& \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) }
\end{eqnarray}
which is equal to the likelihood $\mathcal{L}$. Introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L}(\theta; {\rm data}) &=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:app_main}
\end{eqnarray}
where we define the denominator as $d$ for simplicity in future derivations.

\subsection{Supernova Selection Effects}
\label{app:selection2}
We assume that our selection effects can be reasonably well encapsulated by independent functions of (true) apparent magnitude and redshift, such that $P(S|{\rm data}, \theta) = P(S|z) P(S|m_B)$. Our denominator then becomes
\begin{equation}
d = \int d\hat{z} \, d\hat{m}_B \, dz \, dm_B P(S|z) P(S|m_B) P(\hat{z}|z) P(\hat{m}_B|m_B) P(z, m_B | \theta), \label{eq:sel}
\end{equation}
where for simplicity we have not written out all the integrals that do not interact with the selection effects explicitly. Due to our assumed perfect measurement of redshift, $P(\hat{z}|z) = \delta(\hat{z} - z)$. $P(\hat{m}_B | m_B)$ is a Gaussian due to our Gaussian model of summary statistics, and $m_B$, $x_1$, $c$, and can be analytically integrated out, collapsing the integral over $\hat{m}_B$ (which is why they were not included in equation \eqref{eq:sel}). Finally, we can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.

>>>>

One large difference between traditional methods and BHM methods is that we treat selection effects by incorporating selection efficiency into our model, rather than relying on simulation-driven data corrections. We describe the probability that the {\blue possible} events we observe are drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are observed and pass cuts.  To make this extra conditional explicit, we can write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}
As our model so far describes components of a basic likelihood $P({\rm data}|\theta)$, and we wish to formulate a function $P(S|{\rm data},\theta)$ that describes the chance of an event being successfully observed, we rearrange the likelihood in terms of those functions and find
\begin{align}
\mathcal{L}(\theta; {\rm data}) &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
where the denominator represents an integral over all potential data $D$. This is derived in Appendix \ref{app:selection1}. {\blue To evaluate the effect of our selection effects, we need to evaluate this integral. The integral represents our global selection efficiency at a location in parameter space, rather than the probability of our data being selected into our sample.} As $\theta$ represents the vector of all model parameters, and $D$ represents a vector of all observed variables, this is not a trivial integral. Techniques to approximate this integral, such as Monte-Carlo integration or high-dimensional Gaussian processes failed to give tractable posterior surfaces that could be sampled efficiently by Hamiltonian Monte-Carlo, and post-fitting importance sampling failed due to high-dimensionality (a brief dismissal of many months of struggle). We therefore simplify the integral and approximate the selection effects from their full expression in all of $\theta$-space, to apparent magnitude and redshift space independently (not dependent on $x_1$ or $c$), such that the denominator of equation \eqref{eq:main}, denoted now $d$ for simplicity, is given as
\begin{equation}
d = \int  \left[ \int P(S|m_B) P(m_B | z, \theta)\, d m_B \right] P(S|z) P(z|\theta)\, dz, \label{eq:w1}
\end{equation}
where $P(m_B | z, \theta)$ can be expressed by translating the underlying $M_B$, $x_1$, and $c$ population to $m_B$ given cosmological parameters. A full derivation of this can be found in Appendix \ref{app:selection2}.

@@

\subsection{General Selection Effects}
\label{app:selection1}
When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{equation}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{equation}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments that have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{equation}
\mathcal{L}(\theta; {\rm data}) = P({\rm data} | \theta, S). \label{eq:app_like}
\end{equation}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{eqnarray}
P({\rm data} | S , \theta) P(S, \theta) &=& P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &=& \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) } \label{eq:general1}
\end{eqnarray}
which is equal to the likelihood $\mathcal{L}$. Introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L}(\theta; {\rm data}) &=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:app_main}
\end{eqnarray}
where we define the denominator as $d$ for simplicity in future derivations.

\subsection{Supernova Selection Effects}
\label{app:selection2}
{\blue
To turn the generalised equations from the previous sections into selection effects relevant for this model, we first need to write out our latent parameters, as $\theta$ represents only our top level parameters ($\Omega_m$, $w$, $\alpha$, $\beta$, etc). Whilst they can be inserted into the denominator of equation \eqref{eq:general1} to get $P(S|\theta) = \int P(S,L|\theta)\, dL$ and expanded out, we can simply write down our denominator with general latent parameter(s) $L$, as such:
\begin{equation}
d = \iint  P(S|D, L, \theta) P(D | L, \theta) P(L | \theta)\, dD\, dL
\end{equation}
Here we can now move from the generic labels $D$ and $L$ to something specific to our model. In this case, $D$ would represent our observables, $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{m}, \hat{p}\rbrace$, and $L$ our latent parameters $\lbrace m_B, x_1, c, z\rbrace.$ However, as we assume that our selection effects can be sufficiently encapsulated by true apparent magnitude $m_B$ and redshift $z$, we will not write out the other parameters for simplicity. If we did, we'd simply find them integrated to unity. Formally, we assume that $P(S|D, L, \theta) = P(S|z) P(S|m_B)$.}
Our denominator then becomes
\begin{equation}
d = \int P(S|z) P(S|m_B) P(\hat{z}|z) P(\hat{m}_B|m_B) P(z, m_B | \theta)\, d\hat{z} \, d\hat{m}_B \, dz \, dm_B.  \label{eq:sel} 
\end{equation}
Due to our assumed perfect measurement of redshift, $P(\hat{z}|z) = \delta(\hat{z} - z)$. $P(\hat{m}_B | m_B)$ is a Gaussian due to our Gaussian model of summary statistics, and can be analytically integrated out, collapsing the integral over $\hat{m}_B$. Finally, we can express $P(z, m_B | \theta)$ as  $P(m_B | z, \theta) P(z | \theta)$, where the first term requires us to calculate the magnitude distribution of our underlying population at a given redshift, and the second term is dependent on survey geometry and supernovae rates. We can thus state
\begin{equation}
d = \int \left[ \int P(S|m_B) P(m_B | z, \theta)\, dm_B \right] P(S|z)P(z|\theta)\, dz.
\end{equation}
By assuming that the distribution $P(S|z)P(z|\theta)$ is well sampled by the observed supernova redshifts, we can approximate the integral over redshift by evaluating
\begin{equation}
\int P(S|m_B) P(m_B | z, \theta)\, dm_B \label{eq:selint}
\end{equation}
for each supernova in the dataset -- i.e. Monte Carlo integration with assumed perfect importance sampling.
